---
title: "Final Project"
author: "Yasmin Guadarrama, Harsh Mishrikotkar, Shivanshi Shukla"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load files}

###############################################
# STAT 385 Final Project
# Tetuan City Power Consumption Analysis
###############################################

library(tidyverse)
library(lubridate)
library(GGally)

#--------------------------------------------------
# 1. Load and Inspect Data
#--------------------------------------------------

#setwd("C:/Users/yasmi/OneDrive/Desktop/Final Project")

data <- read_csv("Tetuan City power consumption.csv")

glimpse(data)
summary(data)
sum(duplicated(data))   # No duplicates

#--------------------------------------------------
# 2. Fix DateTime Format (Corrected)
#--------------------------------------------------
# The dataset contains multiple timestamp formats:
# - "1/1/2017 0:00"
# - "01/01/2017 00:00"
# - "2017-01-01 00:00:00"
# parse_date_time() handles all of them.

data <- data %>%
  mutate(DateTime = parse_date_time(
    DateTime,
    orders = c("d/m/Y H:M",
               "d/m/Y H:M:S",
               "Y-m-d H:M:S",
               "m/d/Y H:M",
               "m/d/Y H:M:S")
  ))

# Verify successful parsing
sum(is.na(data$DateTime))   # Should be 0
range(data$DateTime)        # Should show full 2017 range

#--------------------------------------------------
# 3. Clean Column Names (Fix spacing issues)
#--------------------------------------------------

data <- data %>%
  rename(
    Zone1 = `Zone 1 Power Consumption`,
    Zone2 = `Zone 2  Power Consumption`,   # two spaces in original
    Zone3 = `Zone 3  Power Consumption`    # two spaces in original
  )

#--------------------------------------------------
# 4. Exploratory Data Analysis
#--------------------------------------------------

## 4a. Time Series Plot for All Zones
data %>%
  select(DateTime, Zone1, Zone2, Zone3) %>%
  pivot_longer(-DateTime, names_to = "Zone", values_to = "Power") %>%
  ggplot(aes(DateTime, Power, color = Zone)) +
  geom_line(alpha = 0.4) +
  labs(title = "Power Consumption Over Time", y = "Power") +
  theme_minimal()

## 4b. Histograms for All Variables
data %>%
  select(Temperature, Humidity, `Wind Speed`,
         Zone1, Zone2, Zone3) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value") %>%
  ggplot(aes(Value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~ Variable, scales = "free") +
  theme_minimal()

## 4c. Boxplots for Zone Power Consumption
data %>%
  select(Zone1, Zone2, Zone3) %>%
  pivot_longer(everything(), names_to = "Zone", values_to = "Power") %>%
  ggplot(aes(Zone, Power)) +
  geom_boxplot() +
  theme_minimal()

## 4d. Correlation Matrix
data_corr <- data %>%
  select(Temperature, Humidity, `Wind Speed`,
         `general diffuse flows`, `diffuse flows`,
         Zone1, Zone2, Zone3)

ggcorr(data_corr, label = TRUE)

#--------------------------------------------------
# 5. Feature Engineering
#--------------------------------------------------

data <- data %>%
  mutate(
    hour  = hour(DateTime),
    dow   = wday(DateTime, label = TRUE, week_start = 1),
    month = month(DateTime, label = TRUE)
  )

#--------------------------------------------------
# 6. Standardize Numeric Predictors
#--------------------------------------------------

num_vars <- c("Temperature", "Humidity", "Wind Speed",
              "general diffuse flows", "diffuse flows")

data_scaled <- data %>%
  mutate(across(all_of(num_vars), scale))

#--------------------------------------------------
# 7. Time-Based Train/Test Split
#--------------------------------------------------

n <- nrow(data_scaled)
train_end <- floor(0.8 * n)

train <- data_scaled[1:train_end, ]
test  <- data_scaled[(train_end + 1):n, ]

###################################################
# End of Preprocessing
###################################################


```

## Possible questions to address:

- How well can we predict Zone 1, Zone 2, and Zone 3 power consumption using temperature, humidity, wind speed, general diffuse flows, diffuse flows, and time information?
- Which variables are the most important drivers of power demand in each zone?
- Do different zones behave differently, meaning should we build separate models for each zone instead of one combined model?
- Does adding time based features, such as hour of day, day of week, or month, improve prediction accuracy beyond weather-only predictors?
- Which method gives the best tradeoff between accuracy and interpretability for short-term electricity demand forecasting?

