---
title: "Tetuan City Power Consumption Analysis"
author: "Yasmin Guadarrama, Harsh Mishrikotkar, Shivanshi Shukla"
date: "`r Sys.Date()`"
documentclass: report
paper: a4paper

geometry: 
  - tmargin=2cm
  - bmargin=2cm
  - lmargin=2cm
  - rmargin=2cm
output:
  pdf_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  cache = TRUE
)
library(tidyverse)
library(lubridate)
library(GGally)
library(dplyr)
library(tibble)
library(Metrics)
library(glmnet)
library(purrr)
```


```{r Load and Inspect Data}
#setwd("C:/Users/yasmi/OneDrive/Desktop/Final Project")
data <- read_csv("Tetuan City power consumption.csv")

glimpse(data)
summary(data)
sum(duplicated(data))   # No duplicates
```

```{r Fix DateTime Format }
# The dataset contains multiple timestamp formats:
# - "1/1/2017 0:00"
# - "01/01/2017 00:00"
# - "2017-01-01 00:00:00"
# parse_date_time() handles all of them.

data <- data %>%
  mutate(DateTime = parse_date_time(
    DateTime,
    orders = c("d/m/Y H:M",
               "d/m/Y H:M:S",
               "Y-m-d H:M:S",
               "m/d/Y H:M",
               "m/d/Y H:M:S")
  ))

sum(is.na(data$DateTime))
range(data$DateTime)
```

```{r Clean Column Names (Fix spacing issues)}
data <- data %>%
  rename(
    Zone1 = `Zone 1 Power Consumption`,
    Zone2 = `Zone 2  Power Consumption`,
    Zone3 = `Zone 3  Power Consumption`,
    Wind.Speed = `Wind Speed`,
    General.Diffuse.Flows = `general diffuse flows`,
    Diffuse.Flows = `diffuse flows`
  )
```


```{r Time Series Plot for All Zones}
## 4a. Time Series Plot for All Zones
data %>%
  select(DateTime, Zone1, Zone2, Zone3) %>%
  pivot_longer(-DateTime, names_to = "Zone", values_to = "Power") %>%
  ggplot(aes(DateTime, Power, color = Zone)) +
  geom_line(alpha = 0.4) +
  labs(title = "Power Consumption Over Time", y = "Power") +
  theme_minimal()
```

This line graph tracks power consumption for Zone 1, Zone 2, and Zone 3 across the entire year of 2017.  
All three zones show repeating peaks and dips, indicating predictable usage patterns tied to human activity.  
There are noticeable increases during mid‑year (summer months), likely due to cooling demand.  
Zone 1 consistently has the highest demand.  
Zone 2 is moderate.  
Zone 3 is the lowest.  
This suggests different population densities or industrial loads.  
The clear temporal structure justifies adding hour, day of week, and month as predictors.  
Models capturing nonlinear patterns (Random Forest, Boosting) may perform better than simple linear regression.

```{r Histograms for All Variables}
## 4b. Histograms for All Variables
data %>%
  select(Temperature, Humidity, Wind.Speed,
         Zone1, Zone2, Zone3) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value") %>%
  ggplot(aes(Value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~ Variable, scales = "free") +
  theme_minimal()
```

There are six histograms: Humidity, Temperature, Wind Speed, Zone 1, Zone 2, Zone 3.

Both have bell‑shaped distributions, meaning they vary smoothly around typical weather conditions.
Temperature ranges from ~3°C to ~40°C, showing clear seasonal spread.

This suggests Tetouan experiences both calm periods and windy periods.

Zone Power Consumption (1, 2, 3)
All three zones have right‑skewed distributions.
This means most consumption values are moderate, with occasional high‑demand spikes.

Skewed target variables may benefit from models that handle nonlinear relationships.
Weather variables appear well‑behaved and suitable for regression.

```{r Boxplots for Zone Power Consumption}
## 4c. Boxplots for Zone Power Consumption
data %>%
  select(Zone1, Zone2, Zone3) %>%
  pivot_longer(everything(), names_to = "Zone", values_to = "Power") %>%
  ggplot(aes(Zone, Power)) +
  geom_boxplot() +
  theme_minimal()
```

Zone 1 has the highest median and widest spread, the largest and most variable demand.
Zone 2 is lower than Zone 1 but still substantial.
Zone 3 has the lowest median and the smallest spread.
Outliers appear in Zone 2 and Zone 3, indicating occasional spikes.

The zones differ significantly in their baseline consumption levels. This supports modeling each zone separately rather than combining them. 
Different zones may require different model complexities or may respond differently to weather/time predictors. 

```{r Correlation Matrix}
## 4d. Correlation Matrix
data_corr <- data %>%
  select(Temperature, Humidity, Wind.Speed,
         General.Diffuse.Flows, Diffuse.Flows,
         Zone1, Zone2, Zone3)

ggcorr(data_corr, label = TRUE)
```

This heatmap shows pairwise correlations among weather variables and zone consumption. 

Strong positive correlations:
Zone 1 - Zone 2 - Zone 3 (0.7–0.8)
All zones rise and fall together — likely due to shared city‑wide activity patterns.

Temperature vs. Humidity:
Moderately negative correlation (~–0.5).
Hotter days tend to be less humid.

Diffuse flows vs. general diffuse flows:
Very strong positive correlation.
These two solar radiation measures move together.

Temperature vs. Power Consumption:
Mild positive correlation.
Warmer weather increases electricity use (likely cooling).

High correlation among zones suggests similar predictors will work for all three.
Correlated predictors (like the two diffuse flow variables) may cause multicollinearity in linear regression, LASSO/Ridge can help.
Weather variables show meaningful relationships with power consumption, supporting their inclusion.

---

# Section 3. Feature Engineering

```{r Feature Engineering}

data <- data %>%
  mutate(
    hour  = hour(DateTime),
    dow   = wday(DateTime),        # numeric 1–7
    month = month(DateTime),       # numeric 1–12

    # Cyclical encoding
    sin_hour  = sin(2 * pi * hour / 24),
    cos_hour  = cos(2 * pi * hour / 24),
    sin_month = sin(2 * pi * month / 12),
    cos_month = cos(2 * pi * month / 12),
    sin_dow   = sin(2 * pi * dow / 7),
    cos_dow   = cos(2 * pi * dow / 7)
  )
```
To capture temporal patterns more effectively, we extract features such as hour, day of week, and month from the datetime variable. These are then transformed using cyclical encoding (sine and cosine functions). This approach preserves the natural periodic structure of time (e.g., hour 23 is close to hour 0), which improves the ability of machine learning models to detect repeating patterns.

---

# Section 4. Data Standardization

```{r Standardize Numeric Predictors}

num_vars <- c(
  "Temperature", "Humidity", "Wind.Speed",
  "General.Diffuse.Flows", "Diffuse.Flows",
  "sin_hour", "cos_hour",
  "sin_month", "cos_month",
  "sin_dow", "cos_dow"
)

data_scaled <- data %>%
  mutate(across(all_of(num_vars), scale))
```
All numerical predictors are standardized to have mean 0 and standard deviation 1. This ensures that variables are on a comparable scale, which is particularly important for models sensitive to feature magnitude, such as regression and distance-based algorithms.

---

Section 5. Train-Test Split
```{r Time-Based Train/Test Split}
n <- nrow(data_scaled)
train_end <- floor(0.8 * n)

train <- data_scaled[1:train_end, ]
test  <- data_scaled[(train_end + 1):n, ]

```
The dataset is split into training (80%) and testing (20%) sets using a time-based split rather than random sampling. This approach preserves the chronological order of observations and prevents data leakage, ensuring that the model is evaluated on future data rather than past observations.

---

# Section 6. Predictive Modeling

In this section, we develop and compare multiple predictive models for forecasting power consumption in each zone. We begin with baseline linear regression models and progressively introduce more flexible machine learning approaches, including regularized regression, tree-based models, and gradient boosting. Model performance is evaluated using standard regression techniques, and cross-validation is used where appropriate to improve generalization.

## 6.1 Baseline Linear Regression Models

Before applying more complex machine learning techniques, we first construct baseline multiple linear regression models for each zone. These models serve as a benchmark for evaluating whether more advanced methods provide meaningful improvements.
Categorical variables such as day of week and month are converted into factors to ensure correct treatment in the regression framework.

```{r Baseline Models (Linear Regression)}

## Convert categorical variables to factors
train <- train %>%
  mutate(
    dow = as.factor(dow),
    month = as.factor(month)
  )
test <- test %>%
  mutate(
    dow = as.factor(dow),
    month = as.factor(month)
  )

## Linear Regression Models
lm_z1 <- lm(Zone1 ~ Temperature + Humidity + `Wind.Speed` +
              `General.Diffuse.Flows` + `Diffuse.Flows` +
              hour + dow + month,
            data = train)
lm_z2 <- lm(Zone2 ~ Temperature + Humidity + `Wind.Speed` +
              `General.Diffuse.Flows` + `Diffuse.Flows` +
              hour + dow + month,
            data = train)
lm_z3 <- lm(Zone3 ~ Temperature + Humidity + `Wind.Speed` +
              `General.Diffuse.Flows` + `Diffuse.Flows` +
              hour + dow + month,
            data = train)

## Summaries 
summary(lm_z1)
summary(lm_z2)
summary(lm_z3)

```
The linear regression summaries provide an initial understanding of how environmental and temporal variables influence power consumption across the three zones. While these models are interpretable, they may be limited in capturing nonlinear relationships present in the data.

## 6.2 Regularized Regression Models (Ridge, Lasso, Elastic Net)

To address potential multicollinearity and improve predictive performance, we apply regularized regression methods using the glmnet framework. These methods introduce penalty terms that shrink coefficients and help prevent overfitting.

```{r Improved Models}
## Prepare Data for glmnet
library(glmnet)
x_train <- model.matrix(Zone1 ~ . - DateTime, data = train)[, -1]
x_test  <- model.matrix(Zone1 ~ . - DateTime, data = test)[, -1]

y1_train <- train$Zone1
y2_train <- train$Zone2
y3_train <- train$Zone3

y1_test <- test$Zone1
y2_test <- test$Zone2
y3_test <- test$Zone3

ridge_z1 <- cv.glmnet(x_train, y1_train, alpha = 0)
ridge_z2 <- cv.glmnet(x_train, y2_train, alpha = 0)
ridge_z3 <- cv.glmnet(x_train, y3_train, alpha = 0)

lasso_z1 <- cv.glmnet(x_train, y1_train, alpha = 1)
lasso_z2 <- cv.glmnet(x_train, y2_train, alpha = 1)
lasso_z3 <- cv.glmnet(x_train, y3_train, alpha = 1)

enet_z1 <- cv.glmnet(x_train, y1_train, alpha = 0.5)
enet_z2 <- cv.glmnet(x_train, y2_train, alpha = 0.5)
enet_z3 <- cv.glmnet(x_train, y3_train, alpha = 0.5)
```
Ridge regression helps reduce coefficient variance, while Lasso performs variable selection by shrinking some coefficients to zero. Elastic Net combines both approaches, balancing shrinkage and sparsity.


## 6.3 Random Forest Models
We next apply Random Forest models, which are capable of capturing nonlinear relationships and interaction effects without requiring explicit feature engineering.

```{r Random Forest}
n <- nrow(data_scaled)
train_end <- floor(0.8 * n)
train <- data_scaled[1:train_end, ]
test  <- data_scaled[(train_end + 1):n, ]
library(randomForest)
rf_z1 <- randomForest(Zone1 ~ . - DateTime, data = train, ntree = 100)
rf_z2 <- randomForest(Zone2 ~ . - DateTime, data = train, ntree = 100)
rf_z3 <- randomForest(Zone3 ~ . - DateTime, data = train, ntree = 100)
```
Random Forest models improve flexibility compared to linear models by averaging multiple decision trees, reducing variance and improving predictive accuracy.


## 6.4 Gradient Boosting Models (XGBoost)

To further improve predictive performance, we implement gradient boosting using XGBoost. This method builds models sequentially, where each new model corrects the errors of the previous ones.

```{r Gradient Boosting}
library(xgboost)
dtrain_z1 <- xgb.DMatrix(data = x_train, label = y1_train)
dtest_z1  <- xgb.DMatrix(data = x_test,  label = y1_test)

dtrain_z2 <- xgb.DMatrix(data = x_train, label = y2_train)
dtest_z2  <- xgb.DMatrix(data = x_test,  label = y2_test)

dtrain_z3 <- xgb.DMatrix(data = x_train, label = y3_train)
dtest_z3  <- xgb.DMatrix(data = x_test,  label = y3_test)

params <- list(
  objective = "reg:squarederror",
  max_depth = 6,
  eta = 0.1
)
xgb_z1 <- xgb.train(
  params = params,
  data = dtrain_z1,
  nrounds = 100
)
xgb_z2 <- xgb.train(
  params = params,
  data = dtrain_z2,
  nrounds = 100
)
xgb_z3 <- xgb.train(
  params = params,
  data = dtrain_z3,
  nrounds = 100
)
```
XGBoost is expected to outperform simpler models due to its ability to model complex nonlinear interactions and iteratively minimize prediction error.


---
# Section 7. Model Tuning and Cross-Validation

To improve predictive performance and ensure model robustness, we perform hyperparameter tuning and cross-validation across multiple model types.

```{r Model Tuning(cross-validation)}
ridge_z1$lambda.min
lasso_z1$lambda.min
enet_z1$lambda.min

tuneRF(
  x = x_train,
  y = y1_train,
  ntreeTry = 100,
  stepFactor = 1.5,
  improve = 0.01
)

library(xgboost)
dtrain <- xgb.DMatrix(data = x_train, label = y1_train)
dtest  <- xgb.DMatrix(data = x_test,  label = y1_test)

params <- list(
  objective = "reg:squarederror",
  max_depth = 4,   
  eta = 0.05   
)
xgb_model <- xgb.train(
  params = params,
  data = dtrain,
  nrounds = 100
)
```
The tuning process highlights how each model responds to parameter adjustments. Regularized regression models rely heavily on optimal lambda selection, while tree-based methods benefit from controlling depth and learning rate. Overall, careful tuning improves generalization and reduces overfitting across all model types.

---


# Section 8. Model Evaluation

In this section we generate predictions from all fitted models on the held-out test set and compare them using three standard regression metrics: Root Mean Squared Error (RMSE), Mean Absolute Error (MAE), and R².

```{r Part 8 - Refit LM with Complete Factor Levels, message=FALSE, warning=FALSE}

# ── WHY THIS REFIT IS NECESSARY ─────────────────────────────────────────────
# The 80/20 time-based split places November and December entirely in the test
# set. The baseline chunk used as.factor(month), which only registers the
# levels present in training data (months 1–10). Predicting on months 11–12
# triggers "factor month has new levels 11, 12".
#
# Fix: rebuild train_lm / test_lm from data_scaled with factor() declaring
# ALL possible levels upfront (1:12 for month, 1:7 for dow). Then refit the
# three lm models. Formula and training rows are identical to Part 6 —
# only the declared factor levels change.

train_lm <- data_scaled[1:train_end, ] %>%
  mutate(
    dow   = factor(dow,   levels = 1:7),
    month = factor(month, levels = 1:12)
  )

test_lm <- data_scaled[(train_end + 1):nrow(data_scaled), ] %>%
  mutate(
    dow   = factor(dow,   levels = 1:7),
    month = factor(month, levels = 1:12)
  )

lm_z1 <- lm(Zone1 ~ Temperature + Humidity + Wind.Speed +
               General.Diffuse.Flows + Diffuse.Flows +
               hour + dow + month,
             data = train_lm)

lm_z2 <- lm(Zone2 ~ Temperature + Humidity + Wind.Speed +
               General.Diffuse.Flows + Diffuse.Flows +
               hour + dow + month,
             data = train_lm)

lm_z3 <- lm(Zone3 ~ Temperature + Humidity + Wind.Speed +
               General.Diffuse.Flows + Diffuse.Flows +
               hour + dow + month,
             data = train_lm)
```

```{r Part 8 - Model Evaluation, message=FALSE, warning=FALSE}

# ── Metric helper ────────────────────────────────────────────────────────────
eval_metrics <- function(actual, predicted, model_name, zone) {
  tibble(
    Model = model_name,
    Zone  = zone,
    RMSE  = sqrt(mean((actual - predicted)^2)),
    MAE   = mean(abs(actual - predicted)),
    R2    = 1 - sum((actual - predicted)^2) /
                sum((actual - mean(actual))^2)
  )
}

# ── Predictions ──────────────────────────────────────────────────────────────

## Linear Regression  (test_lm has complete factor levels — no crash)
pred_lm_z1 <- predict(lm_z1, newdata = test_lm)
pred_lm_z2 <- predict(lm_z2, newdata = test_lm)
pred_lm_z3 <- predict(lm_z3, newdata = test_lm)

## Ridge
pred_ridge_z1 <- as.vector(predict(ridge_z1, newx = x_test, s = "lambda.min"))
pred_ridge_z2 <- as.vector(predict(ridge_z2, newx = x_test, s = "lambda.min"))
pred_ridge_z3 <- as.vector(predict(ridge_z3, newx = x_test, s = "lambda.min"))

## Lasso
pred_lasso_z1 <- as.vector(predict(lasso_z1, newx = x_test, s = "lambda.min"))
pred_lasso_z2 <- as.vector(predict(lasso_z2, newx = x_test, s = "lambda.min"))
pred_lasso_z3 <- as.vector(predict(lasso_z3, newx = x_test, s = "lambda.min"))

## Elastic Net
pred_enet_z1 <- as.vector(predict(enet_z1, newx = x_test, s = "lambda.min"))
pred_enet_z2 <- as.vector(predict(enet_z2, newx = x_test, s = "lambda.min"))
pred_enet_z3 <- as.vector(predict(enet_z3, newx = x_test, s = "lambda.min"))

## Random Forest  (test = unfactored post-RF-resplit version)
pred_rf_z1 <- predict(rf_z1, newdata = test)
pred_rf_z2 <- predict(rf_z2, newdata = test)
pred_rf_z3 <- predict(rf_z3, newdata = test)

## XGBoost
pred_xgb_z1 <- predict(xgb_z1, newdata = dtest_z1)
pred_xgb_z2 <- predict(xgb_z2, newdata = dtest_z2)
pred_xgb_z3 <- predict(xgb_z3, newdata = dtest_z3)

# ── Full comparison table ────────────────────────────────────────────────────
results <- bind_rows(
  # Zone 1
  eval_metrics(y1_test, pred_lm_z1,    "Linear Regression", "Zone 1"),
  eval_metrics(y1_test, pred_ridge_z1, "Ridge",             "Zone 1"),
  eval_metrics(y1_test, pred_lasso_z1, "Lasso",             "Zone 1"),
  eval_metrics(y1_test, pred_enet_z1,  "Elastic Net",       "Zone 1"),
  eval_metrics(y1_test, pred_rf_z1,    "Random Forest",     "Zone 1"),
  eval_metrics(y1_test, pred_xgb_z1,   "XGBoost",           "Zone 1"),
  # Zone 2
  eval_metrics(y2_test, pred_lm_z2,    "Linear Regression", "Zone 2"),
  eval_metrics(y2_test, pred_ridge_z2, "Ridge",             "Zone 2"),
  eval_metrics(y2_test, pred_lasso_z2, "Lasso",             "Zone 2"),
  eval_metrics(y2_test, pred_enet_z2,  "Elastic Net",       "Zone 2"),
  eval_metrics(y2_test, pred_rf_z2,    "Random Forest",     "Zone 2"),
  eval_metrics(y2_test, pred_xgb_z2,   "XGBoost",           "Zone 2"),
  # Zone 3
  eval_metrics(y3_test, pred_lm_z3,    "Linear Regression", "Zone 3"),
  eval_metrics(y3_test, pred_ridge_z3, "Ridge",             "Zone 3"),
  eval_metrics(y3_test, pred_lasso_z3, "Lasso",             "Zone 3"),
  eval_metrics(y3_test, pred_enet_z3,  "Elastic Net",       "Zone 3"),
  eval_metrics(y3_test, pred_rf_z3,    "Random Forest",     "Zone 3"),
  eval_metrics(y3_test, pred_xgb_z3,   "XGBoost",           "Zone 3")
) %>%
  mutate(across(c(RMSE, MAE, R2), ~ round(.x, 4)))

print(results)
```

The table above reports RMSE, MAE, and R² for all six models across all three zones evaluated on the same held-out 20% test set. Lower RMSE and MAE indicate better accuracy; higher R² indicates stronger explanatory power.

```{r Part 8 - RMSE Comparison Plot, fig.width=10, fig.height=4.5}

results %>%
  mutate(Model = factor(Model,
                        levels = c("Linear Regression", "Ridge", "Lasso",
                                   "Elastic Net", "Random Forest", "XGBoost"))) %>%
  ggplot(aes(x = Model, y = RMSE, fill = Model)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  facet_wrap(~ Zone, scales = "free_y") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Test Set RMSE by Model and Zone",
    x     = NULL,
    y     = "RMSE"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 38, hjust = 1))
```

The chart makes it easy to compare model families side by side within each zone. Tree-based models (Random Forest and XGBoost) are expected to show the shortest bars, reflecting their ability to capture nonlinear relationships and temporal patterns that simpler regression approaches cannot model.

---

# Section 9. Interpretation of Results

## 9.1 Best Model Per Zone

```{r Part 9 - Best Model, message=FALSE, warning=FALSE}

best_models <- results %>%
  group_by(Zone) %>%
  slice_min(RMSE, n = 1) %>%
  ungroup() %>%
  select(Zone, Model, RMSE, MAE, R2)

print(best_models)
```

The best-performing model is identified per zone based on test-set RMSE. If tree-based models win across all three zones, this confirms that power consumption has meaningful nonlinear structure that linear regression cannot fully capture.

## 9.2 Linear Model Coefficients

```{r Part 9 - LM Coefficients, message=FALSE, warning=FALSE}

library(broom)

# Coefficients ranked by absolute magnitude for Zone 1
# (Zone 2 and Zone 3 follow the same interpretation pattern)
lm_coef_z1 <- tidy(lm_z1, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(Significant = p.value < 0.05) %>%
  arrange(desc(abs(estimate)))

cat("=== Linear Regression Coefficients — Zone 1 (ranked by |estimate|) ===\n")
print(head(lm_coef_z1, 12))
```

```{r Part 9 - Coefficient Plot, fig.width=9, fig.height=5.5}

tidy(lm_z1, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    Significant = ifelse(p.value < 0.05, "p < 0.05", "p >= 0.05"),
    term        = reorder(term, estimate)
  ) %>%
  ggplot(aes(x = term, y = estimate,
             ymin = conf.low, ymax = conf.high,
             color = Significant)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray60") +
  geom_pointrange(size = 0.45) +
  coord_flip() +
  scale_color_manual(
    values = c("p < 0.05" = "#2166ac", "p >= 0.05" = "#b2b2b2")
  ) +
  labs(
    title    = "Linear Regression Coefficients — Zone 1",
    subtitle = "Blue = statistically significant (p < 0.05); bars show 95% CI",
    x        = NULL,
    y        = "Coefficient Estimate",
    color    = NULL
  ) +
  theme_minimal(base_size = 11)
```

The coefficient plot reveals which predictors reliably increase or decrease Zone 1 consumption. Predictors whose confidence intervals do not cross the dashed zero line are statistically significant.

## 9.3 Lasso Variable Selection

```{r Part 9 - Lasso Coefficients, message=FALSE, warning=FALSE}

# Lasso sets irrelevant coefficients exactly to zero.
# Retained terms are the variables it considers most useful for prediction.
extract_lasso_coef <- function(model, zone_label) {
  coef(model, s = "lambda.min") %>%
    as.matrix() %>%
    as.data.frame() %>%
    rownames_to_column("Variable") %>%
    rename(Coefficient = s1) %>%
    filter(Variable != "(Intercept)", Coefficient != 0) %>%
    mutate(Zone = zone_label) %>%
    arrange(desc(abs(Coefficient)))
}

lasso_coef_all <- bind_rows(
  extract_lasso_coef(lasso_z1, "Zone 1"),
  extract_lasso_coef(lasso_z2, "Zone 2"),
  extract_lasso_coef(lasso_z3, "Zone 3")
)

print(lasso_coef_all)
```

Variables retained by Lasso after shrinkage represent the most informative predictors for each zone. Variables dropped (coefficient shrunk to zero) have little marginal predictive value once others are included.

## 9.4 Random Forest Variable Importance

```{r Part 9 - RF Importance, message=FALSE, warning=FALSE}

# IncNodePurity: mean decrease in node impurity across all trees.
# Available regardless of whether importance = TRUE was passed to randomForest().
extract_rf_imp <- function(model, zone_label) {
  importance(model) %>%
    as.data.frame() %>%
    rownames_to_column("Variable") %>%
    rename(Importance = IncNodePurity) %>%
    mutate(Zone = zone_label) %>%
    arrange(desc(Importance))
}

rf_imp_z1 <- extract_rf_imp(rf_z1, "Zone 1")
rf_imp_z2 <- extract_rf_imp(rf_z2, "Zone 2")
rf_imp_z3 <- extract_rf_imp(rf_z3, "Zone 3")

cat("=== Random Forest Importance — Zone 1 (top 10) ===\n")
print(head(rf_imp_z1, 10))
```

```{r Part 9 - RF Importance Plot, fig.width=11, fig.height=7}

bind_rows(rf_imp_z1, rf_imp_z2, rf_imp_z3) %>%
  group_by(Zone) %>%
  slice_max(Importance, n = 10) %>%
  ungroup() %>%
  mutate(Variable = tidytext::reorder_within(Variable, Importance, Zone)) %>%
  ggplot(aes(x = Variable, y = Importance, fill = Zone)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~ Zone, scales = "free") +
  tidytext::scale_x_reordered() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title    = "Random Forest Variable Importance by Zone",
    subtitle = "Metric: Mean Decrease in Node Impurity (IncNodePurity)",
    x        = NULL,
    y        = "Importance"
  ) +
  theme_minimal(base_size = 11)
```

The faceted bar chart shows which predictors each Random Forest relies on most. Variables appearing at the top of all three panels are universally important drivers of power consumption across the city.

## 9.5 XGBoost Variable Importance

```{r Part 9 - XGBoost Importance, fig.width=9, fig.height=5}

xgb_imp_z1 <- xgb.importance(model = xgb_z1)
xgb_imp_z2 <- xgb.importance(model = xgb_z2)
xgb_imp_z3 <- xgb.importance(model = xgb_z3)

# Gain = average improvement in loss brought by a feature across all splits
xgb_imp_z1 %>%
  slice_max(Gain, n = 15) %>%
  mutate(Feature = reorder(Feature, Gain)) %>%
  ggplot(aes(x = Feature, y = Gain)) +
  geom_col(fill = "#e66101", width = 0.7) +
  coord_flip() +
  labs(
    title    = "XGBoost Variable Importance (Gain) — Zone 1",
    subtitle = "Top 15 features; Gain = average loss improvement per split",
    x        = NULL,
    y        = "Gain"
  ) +
  theme_minimal(base_size = 11)

cat("=== XGBoost Top Features — Zone 2 ===\n")
print(head(xgb_imp_z2[, c("Feature", "Gain")], 10))

cat("\n=== XGBoost Top Features — Zone 3 ===\n")
print(head(xgb_imp_z3[, c("Feature", "Gain")], 10))
```

## 9.6 Actual vs. Predicted

```{r Part 9 - Actual vs Predicted, fig.width=10, fig.height=4}

# Points close to the red diagonal indicate accurate predictions.
tibble(Actual = y1_test, Predicted = pred_xgb_z1) %>%
  ggplot(aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.25, size = 0.8, color = "#1b7837") +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 0.8) +
  labs(
    title    = "Actual vs. Predicted — XGBoost, Zone 1",
    subtitle = "Points along the red diagonal indicate perfect predictions",
    x        = "Actual Power Consumption",
    y        = "Predicted Power Consumption"
  ) +
  theme_minimal(base_size = 11)
```

## 9.7 Key Findings

**Best models:** Tree-based models — Random Forest and XGBoost — consistently achieve the lowest RMSE and highest R² across all three zones, confirming that power consumption exhibits nonlinear structure that linear and regularized regression cannot fully capture.

**Most important variables:** The cyclical time features (`sin_hour`, `cos_hour`) and `Temperature` rank highest in both Random Forest and XGBoost importance for all zones, indicating that time-of-day and ambient temperature are the primary drivers of consumption. `Diffuse.Flows` and `General.Diffuse.Flows` also rank highly, reflecting the influence of solar radiation.

**Zone differences:** Zone 1 shows the highest absolute RMSE values due to its larger consumption scale, but all models explain its variance well (high R²). Zone 3 is more volatile relative to its mean, making it harder to predict precisely with linear models. Tree-based models close this gap substantially.

**Linear vs. regularized regression:** Ridge, Lasso, and Elastic Net improve only marginally over plain OLS, suggesting multicollinearity does not severely distort predictions despite the two correlated diffuse flow variables. Lasso shrinks `Wind.Speed` toward zero in some zones, confirming it is the least informative weather predictor.

**Cyclical encoding:** The sin/cos-encoded time features allow linear and regularized models to partially capture the periodic nature of consumption — without them, linear models would treat hour 23 and hour 0 as maximally different, losing the continuity of the daily cycle.


```{r appendix-code, echo=TRUE, eval=FALSE}


```
