---
title: "Tetuan City Power Consumption Analysis"
author: "Yasmin Guadarrama, Harsh Mishrikotkar, Shivanshi Shukla"
date: "`r Sys.Date()`"
documentclass: report
paper: a4paper

geometry: 
  - tmargin=2cm
  - bmargin=2cm
  - lmargin=2cm
  - rmargin=2cm
output:
  pdf_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      cache = TRUE)
library(tidyverse)
library(lubridate)
library(GGally)
library(dplyr)
library(tibble)
library(Metrics)
```

# Section 1: Data Setup, Cleaning and Processing

```{r Load and Inspect Data}
#setwd("C:/Users/yasmi/OneDrive/Desktop/Final Project")
data <- read_csv("Tetuan City power consumption.csv")

glimpse(data)
summary(data)
sum(duplicated(data))   # No duplicates
```

```{r Fix DateTime Format }
# The dataset contains multiple timestamp formats:
# - "1/1/2017 0:00"
# - "01/01/2017 00:00"
# - "2017-01-01 00:00:00"
# parse_date_time() handles all of them.

data <- data %>%
  mutate(DateTime = parse_date_time(
    DateTime,
    orders = c("d/m/Y H:M",
               "d/m/Y H:M:S",
               "Y-m-d H:M:S",
               "m/d/Y H:M",
               "m/d/Y H:M:S")
  ))

sum(is.na(data$DateTime))
range(data$DateTime)
```

```{r Clean Column Names (Fix spacing issues)}
data <- data %>%
  rename(
    Zone1 = `Zone 1 Power Consumption`,
    Zone2 = `Zone 2  Power Consumption`,
    Zone3 = `Zone 3  Power Consumption`,
    Wind.Speed = `Wind Speed`,
    General.Diffuse.Flows = `general diffuse flows`,
    Diffuse.Flows = `diffuse flows`
  )
```

```{r Exploratory Data Analysis}
```

```{r Time Series Plot for All Zones}
## 4a. Time Series Plot for All Zones
data %>%
  select(DateTime, Zone1, Zone2, Zone3) %>%
  pivot_longer(-DateTime, names_to = "Zone", values_to = "Power") %>%
  ggplot(aes(DateTime, Power, color = Zone)) +
  geom_line(alpha = 0.4) +
  labs(title = "Power Consumption Over Time", y = "Power") +
  theme_minimal()
```

This line graph tracks power consumption for Zone 1, Zone 2, and Zone 3 across the entire year of 2017.  
All three zones show repeating peaks and dips, indicating predictable usage patterns tied to human activity.  
There are noticeable increases during mid‑year (summer months), likely due to cooling demand.  
Zone 1 consistently has the highest demand.  
Zone 2 is moderate.  
Zone 3 is the lowest.  
This suggests different population densities or industrial loads.  
The clear temporal structure justifies adding hour, day of week, and month as predictors.  
Models capturing nonlinear patterns (Random Forest, Boosting) may perform better than simple linear regression.

```{r Histograms for All Variables}
## 4b. Histograms for All Variables
data %>%
  select(Temperature, Humidity, Wind.Speed,
         Zone1, Zone2, Zone3) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value") %>%
  ggplot(aes(Value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~ Variable, scales = "free") +
  theme_minimal()
```

There are six histograms: Humidity, Temperature, Wind Speed, Zone 1, Zone 2, Zone 3.

Both have bell‑shaped distributions, meaning they vary smoothly around typical weather conditions.
Temperature ranges from ~3°C to ~40°C, showing clear seasonal spread.

This suggests Tetouan experiences both calm periods and windy periods.

Zone Power Consumption (1, 2, 3)
All three zones have right‑skewed distributions.
This means most consumption values are moderate, with occasional high‑demand spikes.

Skewed target variables may benefit from models that handle nonlinear relationships.
Weather variables appear well‑behaved and suitable for regression.

```{r Boxplots for Zone Power Consumption}
## 4c. Boxplots for Zone Power Consumption
data %>%
  select(Zone1, Zone2, Zone3) %>%
  pivot_longer(everything(), names_to = "Zone", values_to = "Power") %>%
  ggplot(aes(Zone, Power)) +
  geom_boxplot() +
  theme_minimal()
```

Zone 1 has the highest median and widest spread, the largest and most variable demand.
Zone 2 is lower than Zone 1 but still substantial.
Zone 3 has the lowest median and the smallest spread.
Outliers appear in Zone 2 and Zone 3, indicating occasional spikes.

The zones differ significantly in their baseline consumption levels. This supports modeling each zone separately rather than combining them. 
Different zones may require different model complexities or may respond differently to weather/time predictors. 

```{r Correlation Matrix}
## 4d. Correlation Matrix
data_corr <- data %>%
  select(Temperature, Humidity, Wind.Speed,
         General.Diffuse.Flows, Diffuse.Flows,
         Zone1, Zone2, Zone3)

ggcorr(data_corr, label = TRUE)
```

This heatmap shows pairwise correlations among weather variables and zone consumption. 

Strong positive correlations:
Zone 1 - Zone 2 - Zone 3 (0.7–0.8)
All zones rise and fall together - likely due to shared city‑wide activity patterns.

Temperature vs. Humidity:
Moderately negative correlation (~–0.5).
Hotter days tend to be less humid.

Diffuse flows vs. general diffuse flows:
Very strong positive correlation.
These two solar radiation measures move together.

Temperature vs. Power Consumption:
Mild positive correlation.
Warmer weather increases electricity use (likely cooling).

High correlation among zones suggests similar predictors will work for all three.
Correlated predictors (like the two diffuse flow variables) may cause multicollinearity in linear regression, LASSO/Ridge can help.
Weather variables show meaningful relationships with power consumption, supporting their inclusion.



# Section 2: Feature Engineering

```{r Feature Engineering}

data <- data %>%
  mutate(
    hour  = hour(DateTime),
    dow   = wday(DateTime),        # numeric 1–7
    month = month(DateTime),       # numeric 1–12

    # Cyclical encoding
    sin_hour  = sin(2 * pi * hour / 24),
    cos_hour  = cos(2 * pi * hour / 24),
    sin_month = sin(2 * pi * month / 12),
    cos_month = cos(2 * pi * month / 12),
    sin_dow   = sin(2 * pi * dow / 7),
    cos_dow   = cos(2 * pi * dow / 7)
  )
```
To capture temporal patterns more effectively, we extract features such as hour, day of week, and month from the datetime variable. These are then transformed using cyclical encoding (sine and cosine functions). This approach preserves the natural periodic structure of time (e.g., hour 23 is close to hour 0), which improves the ability of machine learning models to detect repeating patterns.



# Section 3: Data Standardization

```{r Standardize Numeric Predictors}

num_vars <- c(
  "Temperature", "Humidity", "Wind.Speed",
  "General.Diffuse.Flows", "Diffuse.Flows",
  "sin_hour", "cos_hour",
  "sin_month", "cos_month",
  "sin_dow", "cos_dow"
)

data_scaled <- data %>%
  mutate(across(all_of(num_vars), scale))
```
All numerical predictors are standardized to have mean 0 and standard deviation 1. This ensures that variables are on a comparable scale, which is particularly important for models sensitive to feature magnitude, such as regression and distance-based algorithms.



Section 4. Train-Test Split
```{r Time-Based Train/Test Split}
n <- nrow(data_scaled)
train_end <- floor(0.8 * n)

train <- data_scaled[1:train_end, ]
test  <- data_scaled[(train_end + 1):n, ]

```
The dataset is split into training (80%) and testing (20%) sets using a time-based split rather than random sampling. This approach preserves the chronological order of observations and prevents data leakage, ensuring that the model is evaluated on future data rather than past observations.



# Section 4: Predictive Modeling

In this section, we develop and compare multiple predictive models for forecasting power consumption in each zone. We begin with baseline linear regression models and progressively introduce more flexible machine learning approaches, including regularized regression, tree-based models, and gradient boosting. Model performance is evaluated using standard regression techniques, and cross-validation is used where appropriate to improve generalization.

## 4.1 Baseline Linear Regression Models

Before applying more complex machine learning techniques, we first construct baseline multiple linear regression models for each zone. These models serve as a benchmark for evaluating whether more advanced methods provide meaningful improvements.
Categorical variables such as day of week and month are converted into factors to ensure correct treatment in the regression framework.

```{r Baseline Models (Linear Regression)}

## Convert categorical variables to factors
train <- train %>%
  mutate(
    dow = as.factor(dow),
    month = as.factor(month)
  )
test <- test %>%
  mutate(
    dow = as.factor(dow),
    month = as.factor(month)
  )

## Linear Regression Models
lm_z1 <- lm(Zone1 ~ Temperature + Humidity + `Wind.Speed` +
              `General.Diffuse.Flows` + `Diffuse.Flows` +
              hour + dow + month,
            data = train)
lm_z2 <- lm(Zone2 ~ Temperature + Humidity + `Wind.Speed` +
              `General.Diffuse.Flows` + `Diffuse.Flows` +
              hour + dow + month,
            data = train)
lm_z3 <- lm(Zone3 ~ Temperature + Humidity + `Wind.Speed` +
              `General.Diffuse.Flows` + `Diffuse.Flows` +
              hour + dow + month,
            data = train)

## Summaries 
summary(lm_z1)
summary(lm_z2)
summary(lm_z3)

```
The linear regression summaries provide an initial understanding of how environmental and temporal variables influence power consumption across the three zones. While these models are interpretable, they may be limited in capturing nonlinear relationships present in the data.

## 4.2 Regularized Regression Models (Ridge, Lasso, Elastic Net)

To address potential multicollinearity and improve predictive performance, we apply regularized regression methods using the glmnet framework. These methods introduce penalty terms that shrink coefficients and help prevent overfitting.

```{r Improved Models}
## Prepare Data for glmnet
library(glmnet)
x_train <- model.matrix(Zone1 ~ . - DateTime, data = train)[, -1]
x_test  <- model.matrix(Zone1 ~ . - DateTime, data = test)[, -1]

y1_train <- train$Zone1
y2_train <- train$Zone2
y3_train <- train$Zone3

y1_test <- test$Zone1
y2_test <- test$Zone2
y3_test <- test$Zone3

ridge_z1 <- cv.glmnet(x_train, y1_train, alpha = 0)
ridge_z2 <- cv.glmnet(x_train, y2_train, alpha = 0)
ridge_z3 <- cv.glmnet(x_train, y3_train, alpha = 0)

lasso_z1 <- cv.glmnet(x_train, y1_train, alpha = 1)
lasso_z2 <- cv.glmnet(x_train, y2_train, alpha = 1)
lasso_z3 <- cv.glmnet(x_train, y3_train, alpha = 1)

enet_z1 <- cv.glmnet(x_train, y1_train, alpha = 0.5)
enet_z2 <- cv.glmnet(x_train, y2_train, alpha = 0.5)
enet_z3 <- cv.glmnet(x_train, y3_train, alpha = 0.5)
```
Ridge regression helps reduce coefficient variance, while Lasso performs variable selection by shrinking some coefficients to zero. Elastic Net combines both approaches, balancing shrinkage and sparsity.


## 4.3 Random Forest Models
We next apply Random Forest models, which are capable of capturing nonlinear relationships and interaction effects without requiring explicit feature engineering.

```{r Random Forest}
n <- nrow(data_scaled)
train_end <- floor(0.8 * n)
train <- data_scaled[1:train_end, ]
test  <- data_scaled[(train_end + 1):n, ]
library(randomForest)
rf_z1 <- randomForest(Zone1 ~ . - DateTime, data = train, ntree = 100)
rf_z2 <- randomForest(Zone2 ~ . - DateTime, data = train, ntree = 100)
rf_z3 <- randomForest(Zone3 ~ . - DateTime, data = train, ntree = 100)
```
Random Forest models improve flexibility compared to linear models by averaging multiple decision trees, reducing variance and improving predictive accuracy.


## 4.4 Gradient Boosting Models (XGBoost)

To further improve predictive performance, we implement gradient boosting using XGBoost. This method builds models sequentially, where each new model corrects the errors of the previous ones.

```{r Gradient Boosting}
library(xgboost)
dtrain_z1 <- xgb.DMatrix(data = x_train, label = y1_train)
dtest_z1  <- xgb.DMatrix(data = x_test,  label = y1_test)

dtrain_z2 <- xgb.DMatrix(data = x_train, label = y2_train)
dtest_z2  <- xgb.DMatrix(data = x_test,  label = y2_test)

dtrain_z3 <- xgb.DMatrix(data = x_train, label = y3_train)
dtest_z3  <- xgb.DMatrix(data = x_test,  label = y3_test)

params <- list(
  objective = "reg:squarederror",
  max_depth = 6,
  eta = 0.1
)
xgb_z1 <- xgb.train(
  params = params,
  data = dtrain_z1,
  nrounds = 100
)
xgb_z2 <- xgb.train(
  params = params,
  data = dtrain_z2,
  nrounds = 100
)
xgb_z3 <- xgb.train(
  params = params,
  data = dtrain_z3,
  nrounds = 100
)
```
XGBoost is expected to outperform simpler models due to its ability to model complex nonlinear interactions and iteratively minimize prediction error.



# Section 5: Model Tuning and Cross-Validation

To improve predictive performance and ensure model robustness, we perform hyperparameter tuning and cross-validation across multiple model types.

```{r Model Tuning(cross-validation)}
ridge_z1$lambda.min
lasso_z1$lambda.min
enet_z1$lambda.min

tuneRF(
  x = x_train,
  y = y1_train,
  ntreeTry = 100,
  stepFactor = 1.5,
  improve = 0.01
)

library(xgboost)
dtrain <- xgb.DMatrix(data = x_train, label = y1_train)
dtest  <- xgb.DMatrix(data = x_test,  label = y1_test)

params <- list(
  objective = "reg:squarederror",
  max_depth = 4,   
  eta = 0.05   
)
xgb_model <- xgb.train(
  params = params,
  data = dtrain,
  nrounds = 100
)
```
The tuning process highlights how each model responds to parameter adjustments. Regularized regression models rely heavily on optimal lambda selection, while tree-based methods benefit from controlling depth and learning rate. Overall, careful tuning improves generalization and reduces overfitting across all model types.



# Section 6: Model Evaluation

```{r model-comparison-table}
library(knitr)
library(glmnet)
library(xgboost)

# Rebuild numeric matrices (avoids factor dummy mismatch from earlier chunks)
num_predictors <- c("Temperature", "Humidity", "Wind.Speed",
                    "General.Diffuse.Flows", "Diffuse.Flows",
                    "hour", "dow", "month",
                    "sin_hour", "cos_hour",
                    "sin_month", "cos_month",
                    "sin_dow", "cos_dow")

x_train_s8 <- as.matrix(train[, num_predictors])
x_test_s8  <- as.matrix(test[,  num_predictors])

# Rebuild dtest DMatrix objects (C++ pointers don't persist across chunks)
dtest_z1 <- xgb.DMatrix(data = x_test_s8, label = y1_test)
dtest_z2 <- xgb.DMatrix(data = x_test_s8, label = y2_test)
dtest_z3 <- xgb.DMatrix(data = x_test_s8, label = y3_test)

# Refit glmnet models on clean numeric matrix
ridge_z1 <- cv.glmnet(x_train_s8, y1_train, alpha = 0)
ridge_z2 <- cv.glmnet(x_train_s8, y2_train, alpha = 0)
ridge_z3 <- cv.glmnet(x_train_s8, y3_train, alpha = 0)

lasso_z1 <- cv.glmnet(x_train_s8, y1_train, alpha = 1)
lasso_z2 <- cv.glmnet(x_train_s8, y2_train, alpha = 1)
lasso_z3 <- cv.glmnet(x_train_s8, y3_train, alpha = 1)

enet_z1 <- cv.glmnet(x_train_s8, y1_train, alpha = 0.5)
enet_z2 <- cv.glmnet(x_train_s8, y2_train, alpha = 0.5)
enet_z3 <- cv.glmnet(x_train_s8, y3_train, alpha = 0.5)

# Metrics helper
metrics <- function(actual, pred, model, zone) {
  tibble(
    Model = model, Zone = zone,
    RMSE  = round(sqrt(mean((actual - pred)^2)), 2),
    MAE   = round(mean(abs(actual - pred)), 2),
    R2    = round(1 - sum((actual - pred)^2) / sum((actual - mean(actual))^2), 4)
  )
}

results <- bind_rows(
  # Zone 1
  metrics(y1_test, as.vector(predict(ridge_z1, newx = x_test_s8, s = "lambda.min")), "Ridge",         "Zone 1"),
  metrics(y1_test, as.vector(predict(lasso_z1, newx = x_test_s8, s = "lambda.min")), "Lasso",         "Zone 1"),
  metrics(y1_test, as.vector(predict(enet_z1,  newx = x_test_s8, s = "lambda.min")), "Elastic Net",   "Zone 1"),
  metrics(y1_test, predict(rf_z1,  newdata = test),                                  "Random Forest", "Zone 1"),
  metrics(y1_test, predict(xgb_z1, newdata = dtest_z1),                              "XGBoost",       "Zone 1"),

  # Zone 2
  metrics(y2_test, as.vector(predict(ridge_z2, newx = x_test_s8, s = "lambda.min")), "Ridge",         "Zone 2"),
  metrics(y2_test, as.vector(predict(lasso_z2, newx = x_test_s8, s = "lambda.min")), "Lasso",         "Zone 2"),
  metrics(y2_test, as.vector(predict(enet_z2,  newx = x_test_s8, s = "lambda.min")), "Elastic Net",   "Zone 2"),
  metrics(y2_test, predict(rf_z2,  newdata = test),                                  "Random Forest", "Zone 2"),
  metrics(y2_test, predict(xgb_z2, newdata = dtest_z2),                              "XGBoost",       "Zone 2"),

  # Zone 3
  metrics(y3_test, as.vector(predict(ridge_z3, newx = x_test_s8, s = "lambda.min")), "Ridge",         "Zone 3"),
  metrics(y3_test, as.vector(predict(lasso_z3, newx = x_test_s8, s = "lambda.min")), "Lasso",         "Zone 3"),
  metrics(y3_test, as.vector(predict(enet_z3,  newx = x_test_s8, s = "lambda.min")), "Elastic Net",   "Zone 3"),
  metrics(y3_test, predict(rf_z3,  newdata = test),                                  "Random Forest", "Zone 3"),
  metrics(y3_test, predict(xgb_z3, newdata = dtest_z3),                              "XGBoost",       "Zone 3")
)


```

The performance of all models was evaluated using the same test dataset for each zone. The primary metrics used for comparison were Root Mean Squared Error (RMSE), Mean Absolute Error (MAE), and R-squared (R²). Lower RMSE and MAE indicate better predictive accuracy, while higher R² indicates better explanatory power.

## Zone 1


```{r zone1-table, echo=FALSE}
kable(
  results %>% filter(Zone == "Zone 1") %>% select(-Zone) %>% arrange(RMSE),
  caption = "Model Comparison - Zone 1",
  col.names = c("Model", "RMSE", "MAE", "R²")
)
```
For Zone 1, the regularized linear models performed the best. The Elastic Net model achieved the lowest RMSE (3124.85) and highest R² (0.7432), closely followed by the Lasso model. Ridge regression performed moderately worse, while the Random Forest model showed a noticeable drop in performance. The XGBoost model performed the worst, with a negative R², indicating that it performed worse than simply predicting the mean.

Overall, the results suggest that the relationship between predictors and power consumption in Zone 1 is relatively well-captured by linear models with regularization. The minimal difference between Elastic Net and Lasso indicates that feature selection and shrinkage both contribute similarly in this case.

## Zone 2

```{r zone2-table, echo=FALSE}
kable(
  results %>% filter(Zone == "Zone 2") %>% select(-Zone) %>% arrange(RMSE),
  caption = "Model Comparison - Zone 2",
  col.names = c("Model", "RMSE", "MAE", "R²")
)
```

The results for Zone 2 differ significantly from Zone 1. The Random Forest model achieved the best performance, with the lowest RMSE (5061.52) and the only positive R² (0.1481). All linear and regularized regression models produced negative R² values, indicating poor predictive performance.

XGBoost performed extremely poorly, with a very high RMSE and strongly negative R², suggesting either poor tuning or that the model failed to capture the underlying structure of the data.

These results imply that the relationship between predictors and power consumption in Zone 2 is likely nonlinear and not well captured by linear models. However, even the best-performing model (Random Forest) shows relatively weak predictive power, indicating that important explanatory variables may be missing.

## Zone 3

```{r zone3-table, echo=FALSE}
kable(
  results %>% filter(Zone == "Zone 3") %>% select(-Zone) %>% arrange(RMSE),
  caption = "Model Comparison - Zone 3",
  col.names = c("Model", "RMSE", "MAE", "R²")
)

```

All models performed poorly for Zone 3, with every model producing negative R² values. The XGBoost model achieved the lowest RMSE (4792.20), but still had a negative R², indicating that its predictions were worse than a simple mean-based model.

Random Forest and Ridge regression followed closely behind, but none of the models demonstrated meaningful predictive capability. Elastic Net and Lasso performed especially poorly, with very large errors and extremely negative R² values.

These results suggest that the current feature set fails to adequately explain power consumption in Zone 3. This could be due to missing variables, higher noise levels, or more complex dynamics that are not captured by the models used.

## Overall Comparison

Across all three zones, model performance varies substantially:

- Zone 1: Best modeled by regularized linear models (Elastic Net / Lasso)
- Zone 2: Best modeled by Random Forest, suggesting nonlinear relationships
- Zone 3: No model performs well, indicating a lack of predictive structure in the current data

A key takeaway is that no single model performs best across all zones, reinforcing the decision to model each zone separately.

# Section 7: Interpretation of Results

## 7.1 Importance of Model Choice

The strong performance of Elastic Net and Lasso in Zone 1 suggests that the relationship between predictors and power consumption is largely linear, with some redundancy among predictors making regularization beneficial.

In contrast, the superior performance of Random Forest in Zone 2 indicates the presence of nonlinear relationships or interactions between variables, which linear models fail to capture.

## 7.2 Limitations of the Current Feature Set

The poor performance across Zone 2 and especially Zone 3 suggests that the available predictors (temperature, humidity, wind speed, and flow variables) are not sufficient to fully explain power consumption.

Important missing factors may include human activity patterns, economic behavior, and more detailed temporal effects such as seasonality.

## 7.3 Model Complexity Does Not Guarantee Better Performance

Despite being more advanced, XGBoost performed poorly across all zones. This suggests that either the model was not properly tuned or that the dataset does not support highly complex models.

This reinforces that more complex models are not inherently better.

## 7.4 Implications for Prediction

Zone 1 predictions can be considered reasonably reliable using Elastic Net or Lasso. Zone 2 predictions are weaker but still somewhat usable. Zone 3 predictions are unreliable and should not be used without improving the model or the data.

## 7.5 Recommendations for Improvement

To improve model performance, the following steps are recommended:

- Add more relevant features such as calendar effects or lagged variables
- Explore time-series modeling approaches
- Perform more thorough hyperparameter tuning
- Investigate potential data quality issues


# Section 8: Project Workflow and Contributions

The table below summarizes the workflow of the project along with the primary contributor responsible for each stage.

| Step | Stage                          | Description                                                                 | Responsible |
|------|--------------------------------|-----------------------------------------------------------------------------|-------------|
| 1    | Data Loading & Inspection      | Imported dataset, checked structure, handled data types and missing values | YG     |
| 2    | Exploratory Data Analysis      | Generated summary statistics, visualizations, and identified trends        | YG      |
| 3    | Feature Engineering           | Created time-based and cyclical features for modeling                      | YG      |
| 4    | Data Splitting                | Performed time-based train/test split to prevent data leakage              | YG      |
| 5    | Baseline Modeling             | Built initial linear regression models for each zone                       | SS      |
| 6    | Advanced Modeling             | Implemented Ridge, Lasso, Elastic Net, Random Forest, and XGBoost models   | SS      |
| 7    | Model Tuning                  | Applied cross-validation and selected optimal hyperparameters              | SS      |
| 8    | Model Evaluation              | Compared models using RMSE, MAE, and R² on the test dataset                | HM      |
| 9    | Interpretation & Analysis     | Interpreted results and identified key findings and limitations            | HM      |
| 10    | Final Knitting and Report Preparation     | Created the final Knitted pdf and report finalization            | HM        |


# Appendix
```{r appendix-code, echo=TRUE, eval=FALSE}
# Appendix: Full R Code
# ==============================================
# Tetuan City Power Consumption Analysis
# Yasmin Guadarrama, Harsh Mishrikotkar, Shivanshi Shukla
# ==============================================

# -- Setup ---------------------------------------------------------------------

# knitr::opts_chunk$set(echo = FALSE,
                      cache = TRUE)
library(tidyverse)
library(lubridate)
library(GGally)
library(dplyr)
library(tibble)
library(Metrics)


# -- Section 1: Data Setup, Cleaning and Processing ----------------------------

# Load and Inspect Data
#setwd("C:/Users/yasmi/OneDrive/Desktop/Final Project")
data <- read_csv("Tetuan City power consumption.csv")

glimpse(data)
summary(data)
sum(duplicated(data))   # No duplicates

# Fix DateTime Format
# The dataset contains multiple timestamp formats:
# - "1/1/2017 0:00"
# - "01/01/2017 00:00"
# - "2017-01-01 00:00:00"
# parse_date_time() handles all of them.

data <- data %>%
  mutate(DateTime = parse_date_time(
    DateTime,
    orders = c("d/m/Y H:M",
               "d/m/Y H:M:S",
               "Y-m-d H:M:S",
               "m/d/Y H:M",
               "m/d/Y H:M:S")
  ))

sum(is.na(data$DateTime))
range(data$DateTime)

# Clean Column Names (Fix spacing issues)
data <- data %>%
  rename(
    Zone1 = `Zone 1 Power Consumption`,
    Zone2 = `Zone 2  Power Consumption`,
    Zone3 = `Zone 3  Power Consumption`,
    Wind.Speed = `Wind Speed`,
    General.Diffuse.Flows = `general diffuse flows`,
    Diffuse.Flows = `diffuse flows`
  )


# -- Exploratory Data Analysis --------------------------------------------------

# Time Series Plot for All Zones
data %>%
  select(DateTime, Zone1, Zone2, Zone3) %>%
  pivot_longer(-DateTime, names_to = "Zone", values_to = "Power") %>%
  ggplot(aes(DateTime, Power, color = Zone)) +
  geom_line(alpha = 0.4) +
  labs(title = "Power Consumption Over Time", y = "Power") +
  theme_minimal()

# Histograms for All Variables
data %>%
  select(Temperature, Humidity, Wind.Speed,
         Zone1, Zone2, Zone3) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value") %>%
  ggplot(aes(Value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~ Variable, scales = "free") +
  theme_minimal()

# Boxplots for Zone Power Consumption
data %>%
  select(Zone1, Zone2, Zone3) %>%
  pivot_longer(everything(), names_to = "Zone", values_to = "Power") %>%
  ggplot(aes(Zone, Power)) +
  geom_boxplot() +
  theme_minimal()

# Correlation Matrix
data_corr <- data %>%
  select(Temperature, Humidity, Wind.Speed,
         General.Diffuse.Flows, Diffuse.Flows,
         Zone1, Zone2, Zone3)

ggcorr(data_corr, label = TRUE)


# -- Section 2: Feature Engineering --------------------------------------------

data <- data %>%
  mutate(
    hour  = hour(DateTime),
    dow   = wday(DateTime),        # numeric 1-7
    month = month(DateTime),       # numeric 1-12

    # Cyclical encoding
    sin_hour  = sin(2 * pi * hour / 24),
    cos_hour  = cos(2 * pi * hour / 24),
    sin_month = sin(2 * pi * month / 12),
    cos_month = cos(2 * pi * month / 12),
    sin_dow   = sin(2 * pi * dow / 7),
    cos_dow   = cos(2 * pi * dow / 7)
  )


# -- Section 3: Data Standardization -------------------------------------------

num_vars <- c(
  "Temperature", "Humidity", "Wind.Speed",
  "General.Diffuse.Flows", "Diffuse.Flows",
  "sin_hour", "cos_hour",
  "sin_month", "cos_month",
  "sin_dow", "cos_dow"
)

data_scaled <- data %>%
  mutate(across(all_of(num_vars), scale))


# -- Section 4 (Pre-Modeling): Time-Based Train/Test Split ---------------------

n <- nrow(data_scaled)
train_end <- floor(0.8 * n)

train <- data_scaled[1:train_end, ]
test  <- data_scaled[(train_end + 1):n, ]


# -- Section 4.1: Baseline Linear Regression Models ----------------------------

# Convert categorical variables to factors
train <- train %>%
  mutate(
    dow = as.factor(dow),
    month = as.factor(month)
  )
test <- test %>%
  mutate(
    dow = as.factor(dow),
    month = as.factor(month)
  )

# Linear Regression Models
lm_z1 <- lm(Zone1 ~ Temperature + Humidity + `Wind.Speed` +
              `General.Diffuse.Flows` + `Diffuse.Flows` +
              hour + dow + month,
            data = train)
lm_z2 <- lm(Zone2 ~ Temperature + Humidity + `Wind.Speed` +
              `General.Diffuse.Flows` + `Diffuse.Flows` +
              hour + dow + month,
            data = train)
lm_z3 <- lm(Zone3 ~ Temperature + Humidity + `Wind.Speed` +
              `General.Diffuse.Flows` + `Diffuse.Flows` +
              hour + dow + month,
            data = train)

# Summaries
summary(lm_z1)
summary(lm_z2)
summary(lm_z3)


# -- Section 4.2: Regularized Regression Models (Ridge, Lasso, Elastic Net) ----

library(glmnet)
x_train <- model.matrix(Zone1 ~ . - DateTime, data = train)[, -1]
x_test  <- model.matrix(Zone1 ~ . - DateTime, data = test)[, -1]

y1_train <- train$Zone1
y2_train <- train$Zone2
y3_train <- train$Zone3

y1_test <- test$Zone1
y2_test <- test$Zone2
y3_test <- test$Zone3

ridge_z1 <- cv.glmnet(x_train, y1_train, alpha = 0)
ridge_z2 <- cv.glmnet(x_train, y2_train, alpha = 0)
ridge_z3 <- cv.glmnet(x_train, y3_train, alpha = 0)

lasso_z1 <- cv.glmnet(x_train, y1_train, alpha = 1)
lasso_z2 <- cv.glmnet(x_train, y2_train, alpha = 1)
lasso_z3 <- cv.glmnet(x_train, y3_train, alpha = 1)

enet_z1 <- cv.glmnet(x_train, y1_train, alpha = 0.5)
enet_z2 <- cv.glmnet(x_train, y2_train, alpha = 0.5)
enet_z3 <- cv.glmnet(x_train, y3_train, alpha = 0.5)


# -- Section 4.3: Random Forest Models -----------------------------------------

n <- nrow(data_scaled)
train_end <- floor(0.8 * n)
train <- data_scaled[1:train_end, ]
test  <- data_scaled[(train_end + 1):n, ]

library(randomForest)
rf_z1 <- randomForest(Zone1 ~ . - DateTime, data = train, ntree = 100)
rf_z2 <- randomForest(Zone2 ~ . - DateTime, data = train, ntree = 100)
rf_z3 <- randomForest(Zone3 ~ . - DateTime, data = train, ntree = 100)


# -- Section 4.4: Gradient Boosting Models (XGBoost) ---------------------------

library(xgboost)
dtrain_z1 <- xgb.DMatrix(data = x_train, label = y1_train)
dtest_z1  <- xgb.DMatrix(data = x_test,  label = y1_test)

dtrain_z2 <- xgb.DMatrix(data = x_train, label = y2_train)
dtest_z2  <- xgb.DMatrix(data = x_test,  label = y2_test)

dtrain_z3 <- xgb.DMatrix(data = x_train, label = y3_train)
dtest_z3  <- xgb.DMatrix(data = x_test,  label = y3_test)

params <- list(
  objective = "reg:squarederror",
  max_depth = 6,
  eta = 0.1
)
xgb_z1 <- xgb.train(
  params = params,
  data = dtrain_z1,
  nrounds = 100
)
xgb_z2 <- xgb.train(
  params = params,
  data = dtrain_z2,
  nrounds = 100
)
xgb_z3 <- xgb.train(
  params = params,
  data = dtrain_z3,
  nrounds = 100
)


# -- Section 5: Model Tuning and Cross-Validation ------------------------------

ridge_z1$lambda.min
lasso_z1$lambda.min
enet_z1$lambda.min

tuneRF(
  x = x_train,
  y = y1_train,
  ntreeTry = 100,
  stepFactor = 1.5,
  improve = 0.01
)

library(xgboost)
dtrain <- xgb.DMatrix(data = x_train, label = y1_train)
dtest  <- xgb.DMatrix(data = x_test,  label = y1_test)

params <- list(
  objective = "reg:squarederror",
  max_depth = 4,
  eta = 0.05
)
xgb_model <- xgb.train(
  params = params,
  data = dtrain,
  nrounds = 100
)


# -- Section 6: Model Evaluation -----------------------------------------------

library(knitr)
library(glmnet)
library(xgboost)

# Rebuild numeric matrices (avoids factor dummy mismatch from earlier chunks)
num_predictors <- c("Temperature", "Humidity", "Wind.Speed",
                    "General.Diffuse.Flows", "Diffuse.Flows",
                    "hour", "dow", "month",
                    "sin_hour", "cos_hour",
                    "sin_month", "cos_month",
                    "sin_dow", "cos_dow")

x_train_s8 <- as.matrix(train[, num_predictors])
x_test_s8  <- as.matrix(test[,  num_predictors])

# Rebuild dtest DMatrix objects (C++ pointers don't persist across chunks)
dtest_z1 <- xgb.DMatrix(data = x_test_s8, label = y1_test)
dtest_z2 <- xgb.DMatrix(data = x_test_s8, label = y2_test)
dtest_z3 <- xgb.DMatrix(data = x_test_s8, label = y3_test)

# Refit glmnet models on clean numeric matrix
ridge_z1 <- cv.glmnet(x_train_s8, y1_train, alpha = 0)
ridge_z2 <- cv.glmnet(x_train_s8, y2_train, alpha = 0)
ridge_z3 <- cv.glmnet(x_train_s8, y3_train, alpha = 0)

lasso_z1 <- cv.glmnet(x_train_s8, y1_train, alpha = 1)
lasso_z2 <- cv.glmnet(x_train_s8, y2_train, alpha = 1)
lasso_z3 <- cv.glmnet(x_train_s8, y3_train, alpha = 1)

enet_z1 <- cv.glmnet(x_train_s8, y1_train, alpha = 0.5)
enet_z2 <- cv.glmnet(x_train_s8, y2_train, alpha = 0.5)
enet_z3 <- cv.glmnet(x_train_s8, y3_train, alpha = 0.5)

# Metrics helper
metrics <- function(actual, pred, model, zone) {
  tibble(
    Model = model, Zone = zone,
    RMSE  = round(sqrt(mean((actual - pred)^2)), 2),
    MAE   = round(mean(abs(actual - pred)), 2),
    R2    = round(1 - sum((actual - pred)^2) / sum((actual - mean(actual))^2), 4)
  )
}

results <- bind_rows(
  # Zone 1
  metrics(y1_test, as.vector(predict(ridge_z1, newx = x_test_s8, s = "lambda.min")), "Ridge",         "Zone 1"),
  metrics(y1_test, as.vector(predict(lasso_z1, newx = x_test_s8, s = "lambda.min")), "Lasso",         "Zone 1"),
  metrics(y1_test, as.vector(predict(enet_z1,  newx = x_test_s8, s = "lambda.min")), "Elastic Net",   "Zone 1"),
  metrics(y1_test, predict(rf_z1,  newdata = test),                                  "Random Forest", "Zone 1"),
  metrics(y1_test, predict(xgb_z1, newdata = dtest_z1),                              "XGBoost",       "Zone 1"),

  # Zone 2
  metrics(y2_test, as.vector(predict(ridge_z2, newx = x_test_s8, s = "lambda.min")), "Ridge",         "Zone 2"),
  metrics(y2_test, as.vector(predict(lasso_z2, newx = x_test_s8, s = "lambda.min")), "Lasso",         "Zone 2"),
  metrics(y2_test, as.vector(predict(enet_z2,  newx = x_test_s8, s = "lambda.min")), "Elastic Net",   "Zone 2"),
  metrics(y2_test, predict(rf_z2,  newdata = test),                                  "Random Forest", "Zone 2"),
  metrics(y2_test, predict(xgb_z2, newdata = dtest_z2),                              "XGBoost",       "Zone 2"),

  # Zone 3
  metrics(y3_test, as.vector(predict(ridge_z3, newx = x_test_s8, s = "lambda.min")), "Ridge",         "Zone 3"),
  metrics(y3_test, as.vector(predict(lasso_z3, newx = x_test_s8, s = "lambda.min")), "Lasso",         "Zone 3"),
  metrics(y3_test, as.vector(predict(enet_z3,  newx = x_test_s8, s = "lambda.min")), "Elastic Net",   "Zone 3"),
  metrics(y3_test, predict(rf_z3,  newdata = test),                                  "Random Forest", "Zone 3"),
  metrics(y3_test, predict(xgb_z3, newdata = dtest_z3),                              "XGBoost",       "Zone 3")
)

# Zone 1 Results Table
kable(
  results %>% filter(Zone == "Zone 1") %>% select(-Zone) %>% arrange(RMSE),
  caption = "Model Comparison - Zone 1",
  col.names = c("Model", "RMSE", "MAE", "R²")
)

# Zone 2 Results Table
kable(
  results %>% filter(Zone == "Zone 2") %>% select(-Zone) %>% arrange(RMSE),
  caption = "Model Comparison - Zone 2",
  col.names = c("Model", "RMSE", "MAE", "R²")
)

# Zone 3 Results Table
kable(
  results %>% filter(Zone == "Zone 3") %>% select(-Zone) %>% arrange(RMSE),
  caption = "Model Comparison - Zone 3",
  col.names = c("Model", "RMSE", "MAE", "R²")
)

```